---
title: "Practical statistical inference"
author: "Mike Frank"
date: "10/27/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(BayesFactor)
library(compute.es)
```

# Dataset

In this markdown, we're going to consider statistical inferences in a simple dataset, a subset of the data from:

> Barner, D., Athanasopoulou, A., Chu, J., Lewis, M. L., Marchand, E., Schneider, R. M., Frank, M. C. (in press). A one-year classroom-randomized trial of mental abacus instruction for first- and second-grade students. Journal of Numerical Cognition.

First- and second-grade classrooms were randomly assigned to mental-abacus practice or control. We'll be looking at only the outcome data, and only their scores on a measure of place-value understanding. 

```{r}
d <- read_csv("https://raw.githubusercontent.com/langcog/majic/master/data/processed_data.csv") %>%
  filter(year == "2016") %>%
  select(subid, class_num, grade, group, pvAvg)

d
```

```{r}
d %>%
  group_by(group) %>%
    summarise(pv = mean(pvAvg, na.rm=TRUE), 
              n = n()) 
```

```{r}
ggplot(d, aes(x = pvAvg, fill = group)) + 
  geom_histogram(binwidth = .1, position = "dodge")
```

```{r}
ms <- d %>%
  group_by(group) %>%
  summarise(avg = mean(pvAvg, na.rm=TRUE),
            ci = 1.96 * sd(pvAvg, na.rm=TRUE) / sqrt(sum(!is.na(pvAvg))))

ggplot(ms, 
       aes(x = group, y = avg)) + 
  geom_pointrange(aes(ymin= avg - ci, ymax = avg + ci)) + 
  ylim(0,1)
```



# Statistical Inference

First, let's look at a t-test for the entire dataset.

```{r}
t <- with(d, 
          t.test(pvAvg[group == "MA"], pvAvg[group == "CNTL"]))

# the same as:
# t.test(d$pvAvg[d$group == "MA"], d$pvAvg[d$group == "CNTL"]))

t
```

Let's try a BayesFactor analysis. Documentation [here](http://bayesfactorpcl.r-forge.r-project.org/).

This is a "JZS" test that assumes that the mean is zero under H0, and that H1 is a Cauchy (fat-tailed) distribution with a particular parameter ($\sqrt{2}/2$). 

```{r}
bf <- with(filter(d, !is.na(pvAvg)), 
           ttestBF(x = pvAvg[group == "MA"], y = pvAvg[group == "CNTL"]))

bf
```

Compute effect size.

```{r}
n1 <- sum(!is.na(d$pvAvg[d$group == "MA"]))
n2 <- sum(!is.na(d$pvAvg[d$group == "CNTL"]))
compute.es::tes(t = t$statistic, n.1 = n1, n.2 = n2)
```

# Grade-based analysis

Consider breaking down the data by grade.

```{r}
d %>%
  group_by(group, grade) %>%
    summarise(pv = mean(pvAvg, na.rm=TRUE), 
            n = n()) %>%
  knitr::kable(digits = 2)
```


```{r}
ggplot(d, aes(x = pvAvg, fill = group)) + 
  geom_histogram(binwidth = .2, position = "dodge") +
  facet_wrap(~grade)
```
```{r}
ms <- d %>%
  group_by(grade, group) %>%
  summarise(avg = mean(pvAvg, na.rm=TRUE),
            ci = 1.96 * sd(pvAvg, na.rm=TRUE) / sqrt(sum(!is.na(pvAvg))))

ggplot(ms, 
       aes(x = grade, y = avg, col = group)) + 
  geom_pointrange(aes(ymin = avg - ci, ymax = avg + ci)) +
  geom_line(aes(group = group)) +
  ylim(0,1)
```

```{r}
with(filter(d, grade == "second grade"), 
          t.test(pvAvg[group == "MA"], pvAvg[group == "CNTL"]))
```

```{r}
with(filter(d, grade == "second grade"), 
     ttestBF(x = pvAvg[group == "MA"], y = pvAvg[group == "CNTL"]))
```


```{r}
summary(lm(pvAvg ~ grade * group, data = d))
```


# Conclusions

What we said:

> "Neither age group showed a significant advantage in cognitive abilities or mathematical computation relative to controls, although older children showed some hints of an advantage in learning place-value concepts."
